---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# cryptRopen

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/cryptRopen)](https://CRAN.R-project.org/package=cryptRopen)
[![R-CMD-check](https://github.com/danielrak/cryptRopen/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/danielrak/cryptRopen/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/danielrak/cryptRopen/graph/badge.svg)](https://app.codecov.io/gh/danielrak/cryptRopen)
<!-- badges: end -->

cryptRopen pseudonymizes variables in one or many datasets, driven by
a single Excel mask that lists, per input file, which columns to hash
and which to drop. The transformation is a **salted hash** (MD5 by
default; any algorithm supported by `digest::digest()` works), which
means:

* The same input value paired with the same key always yields the
  same hash. Joins across tables on the pseudonymized columns stay
  possible.
* The hash is one-way: holding the output does not let anyone recover
  the input. Only the correspondence table (kept by the party that
  owns the key) does.

Suitable for GDPR-style pseudonymization. **Not** suitable as a
reversible encryption scheme for confidentiality-bearing data.

## Installation

From CRAN:

``` r
install.packages("cryptRopen")
```

Development version from GitHub:

``` r
# install.packages("pak")
pak::pak("danielrak/cryptRopen")
```

## In-session use: `crypt_data()`

Pseudonymize one data frame already in memory. The correspondence
table is kept in a package-private environment and retrieved with
`get_correspondence_tables()`:

```{r readme-crypt-data}
library(cryptRopen)

pseudonymized <- crypt_data(
  loaded_dataset             = head(mtcars, 3),
  vars_to_encrypt            = "mpg",
  vars_to_remove             = "cyl",
  encryption_key             = "demo-key",
  correspondence_table       = TRUE,
  correspondence_table_label = "demo"
)
pseudonymized

get_correspondence_tables()$tc_crypt_demo
```

## Batch / file-driven use: `crypt_r()`

`crypt_r()` reads an Excel mask describing N input files, dispatches
one `mirai::mirai()` task per row (parallel, non-blocking), writes the
pseudonymized files and a recap xlsx log to disk, and returns a
`cryptR_job` handle. Inspect it with `cryptR_status()` / `summary()` /
`cryptR_results()` and finalize with `cryptR_collect()`.

``` r
job <- crypt_r(
  mask_folder_path  = "path/to/mask",
  mask_file         = "mask.xlsx",
  output_path       = "path/to/output",
  intermediate_path = "path/to/intermediate",
  encryption_key    = "your-salt",
  algorithm         = "md5"
)

cryptR_status(job)
job <- cryptR_collect(job)
```

Each row of the mask declares its own `vars_to_encrypt`, so different
input files can have entirely different column names — the mask is
the contract, not a single schema. Large parquet / CSV inputs are
automatically routed to a streaming engine (`arrow::open_dataset()` +
chunked write) so memory usage stays bounded regardless of input
size.

## Learn more

* `vignette("cryptRopen")` — getting started: the in-session workflow
  with `crypt_vector()` / `crypt_data()`, the difference between
  pseudonymization and encryption, and how to pick an algorithm.
* `vignette("crypt_r-workflow")` — full walkthrough of the
  mask-driven `crypt_r()` workflow: Excel mask layout, async
  inspection, streaming, daemons ownership, recap log.
* `NEWS.md` — change history.

## License

MIT © Daniel Rakotomalala.
