---
title: "Development actions history"
output: html_document
editor_options:
  chunk_output_type: console
---

Development helpers for cryptRopen (post-fusen).

# Common actions

```{r, eval=FALSE}
# Simulate package installation during development
devtools::load_all()

# Regenerate roxygen man/ pages and NAMESPACE from R/ sources
devtools::document()

# Auto-manage Imports/Suggests in DESCRIPTION from NAMESPACE + R/ scan
attachment::att_amend_desc()

# Run tests
devtools::test()

# Full R CMD check (--as-cran adds CRAN policy checks)
devtools::check()
devtools::check(args = c("--as-cran"))
```

# Baseline non-regression fixtures

```{r, eval=FALSE}
# Regenerate the fixtures under tests/baseline/outputs/ after an intentional
# semantic change (keep commits green).
source("tests/baseline/generate_baseline.R")
```

# Publishing (GitHub)

```{r, eval=FALSE}
# Build the tarball
devtools::build()

# Build the pkgdown site locally
pkgdown::build_site()
```

# Notes

This package historically used {fusen} (inflate `dev/flat_*.Rmd` → `R/*.R`).
fusen was retired in April 2026 because the flat templates had drifted from
the generated R sources; `R/*.R` is now the single source of truth. Edit
`R/` directly and run `devtools::document()` to regenerate man pages.
