---
title: "Getting started with cryptRopen"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with cryptRopen}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## What cryptRopen is for

`cryptRopen` pseudonymizes variables inside data frames so that sensitive
values (identifiers, emails, names) can be shared or stored in a
derivative form without exposing the originals. The pseudonymization is
**deterministic**: the same input value paired with the same key always
produces the same output, so joins across tables remain possible on the
pseudonymized columns. A side table — the *correspondence table* —
keeps the mapping between original and pseudonymized values for the
party that holds the key.

Typical use cases: preparing an extract for an analytics team that
should not see customer identifiers; sharing a dataset with an external
partner; feeding a downstream pipeline while keeping the original
identifiers under control.

## Pseudonymization, not encryption

The `crypt` in the package name is historical. The underlying
transformation is a **salted hash**, not reversible encryption:

* A normalized version of each input value is concatenated with a
  user-provided salt (`key`) and hashed with `digest::digest()`.
* The default algorithm is `"md5"`. Any algorithm supported by
  `digest::digest()` (`"sha1"`, `"sha256"`, `"sha512"`, …) is accepted
  via the `algorithm` parameter — see *Algorithm choice* below.
* Holding the output does **not** let anyone recover the input. Only
  the correspondence table (or a rainbow-table-style attack on a weak
  key) does.

Consequence: cryptRopen is suitable for pseudonymization under
regulations such as GDPR, not for confidentiality-bearing encryption.

## When *not* to use cryptRopen

Pseudonymization is one tool among several. Reach for a different
approach when:

* **You need to recover the original value later.** A salted hash is
  one-way. If your workflow must un-pseudonymize at some point, use
  symmetric encryption (e.g. the `safer` or `sodium` packages) and
  manage the key separately.
* **You need formal anonymization, not pseudonymization.** Under
  regulations such as GDPR, pseudonymized data is still personal
  data — the correspondence table is the de facto re-identification
  key. For irreversible anonymization, consider differential privacy,
  k-anonymity, or aggregation.
* **You want to hide value distributions.** A deterministic hash
  preserves the frequency profile of the original column. An
  attacker who knows the rough distribution of the input (e.g. that
  certain customer IDs occur very often) can still spot the most
  frequent hashes. Add random padding or switch to format-preserving
  encryption if that matters for your threat model.
* **You need integrity guarantees on the data, not just on its
  identifiers.** Hashes here are computed *per cell*, not over a
  whole row or file. They do not detect tampering at the dataset
  level — pair them with a separate signature scheme if you need
  that.

## Algorithm choice

The package delegates the actual hashing to `digest::digest()`, so
every algorithm of that function is available via the `algorithm`
parameter. The trade-offs you should know:

* **`"md5"` (default)** — fast and short (32 hex chars). Cryptographic
  collisions exist as theoretical constructions but are negligible at
  practical dataset scales for *pseudonymization*. Recommended unless
  you have a specific reason to step up.
* **`"sha1"`** — 40 hex chars, deprecated in formal cryptography but
  still acceptable for non-adversarial pseudonymization. Rarely a
  better choice than MD5 or SHA-256.
* **`"sha256"`** — 64 hex chars, the modern conservative default if
  you want to be on the safer side. Slower than MD5 but the
  difference is typically below 1 % of total runtime on a real
  pseudonymization job dominated by I/O.
* **`"sha512"`** — 128 hex chars, only useful when downstream
  systems happen to expect a 512-bit hash. The output is twice as
  long for no practical security gain in this context.

Whatever you pick, lock it down in your run and keep the *same*
`encryption_key` across all runs that need to join: changing either
breaks the deterministic property and your tables stop joining on
the pseudonymized columns.

## Installation

```{r install, eval = FALSE}
# install.packages("pak")
pak::pak("danielrak/cryptRopen")
```

```{r setup}
library(cryptRopen)
```

## Hashing a single vector

`crypt_vector()` is the lowest-level entry point. It takes a vector, a
key, and a hash algorithm, and returns a character vector of the same
length. `NA` and empty strings stay `NA`. Input is uppercased, trimmed,
and stripped of internal spaces before hashing, so `"Alice"`,
`"alice "` and `" ALICE "` collapse to the same hash.

```{r crypt-vector-basic}
crypt_vector(
  vector = c("Alice", "Bob", "Charlie"),
  key = "my-secret-salt",
  algo = "md5"
)
```

Determinism and the NA / empty-string rules:

```{r crypt-vector-properties}
# Same value + same key => same hash, every time.
identical(
  crypt_vector("Alice", key = "k", algo = "md5"),
  crypt_vector("alice ", key = "k", algo = "md5")
)

# NAs and empty strings are preserved.
crypt_vector(c("Alice", NA, "", "  "), key = "k", algo = "md5")
```

## Hashing one or several columns of a data frame

`crypt_data()` operates on a data frame and returns a new data frame
with the requested columns replaced by their hashed counterparts
(suffix `_crypt`). Original columns can optionally be dropped
(`vars_to_remove`).

```{r crypt-data-basic}
head(mtcars, 3)

encrypted <- crypt_data(
  loaded_dataset            = head(mtcars, 3),
  vars_to_encrypt           = "mpg",
  vars_to_remove            = "cyl",
  encryption_key            = "demo-key",
  algorithm                 = "md5",
  correspondence_table      = TRUE,
  correspondence_table_label = "demo"
)
encrypted
```

The correspondence table is **not** returned alongside the data frame;
it is stored inside a package-private environment and retrieved via
`get_correspondence_tables()`:

```{r get-tc}
tcs <- get_correspondence_tables()
names(tcs)
tcs$tc_crypt_demo
```

Tables accumulate across calls in the same session. Pass a character
vector to `names` to pull only the ones you need:

```{r get-tc-filter}
get_correspondence_tables(names = "tc_crypt_demo")
```

Pseudonymising multiple columns at once is a matter of extending
`vars_to_encrypt`:

```{r crypt-data-multi}
crypt_data(
  loaded_dataset            = head(mtcars, 3),
  vars_to_encrypt           = c("mpg", "hp"),
  encryption_key            = "demo-key",
  correspondence_table      = TRUE,
  correspondence_table_label = "demo_multi"
)
```

## When you need to batch multiple files

`crypt_data()` is the right tool when you hold one data frame in memory
and want an encrypted version of it. For an industrialized workflow —
several input files specified in an Excel mask, outputs written to
disk, parallel execution, a recap log — see `vignette("crypt_r-workflow")`.

## Related resources

* `?crypt_vector`, `?crypt_data`, `?crypt_r` — the three public
  entry points.
* `?get_correspondence_tables` — retrieval of the correspondence
  tables stored in the package-private environment.
* `NEWS.md` — full change history.
