---
title: "Industrialized workflow with crypt_r()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Industrialized workflow with crypt_r()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`crypt_r()` is the industrialised counterpart of `crypt_data()`. It is
designed for production pipelines that must pseudonymise several input
files at once, write their encrypted versions to disk, and produce a
recap log — all in parallel, without blocking the main R session.

This vignette walks through a complete happy-path example and then
summarises the main options.

```{r setup}
library(cryptRopen)
```

## When to use `crypt_r()` vs. `crypt_data()`

| Need                                                                 | Use              |
| -------------------------------------------------------------------- | ---------------- |
| Pseudonymise one data frame already in memory                        | `crypt_data()`   |
| Pseudonymise N files on disk, described by an Excel mask             | `crypt_r()`      |
| Parallel execution across rows of the mask                           | `crypt_r()`      |
| Streaming read/write for files too large for RAM (parquet, CSV)      | `crypt_r()`      |
| Auto-written xlsx recap log (success, duration, sha256 per row)      | `crypt_r()`      |

## The Excel mask

`crypt_r()` is driven by a single Excel mask that describes what to
encrypt and where. Required columns:

| Column            | Content                                                                   |
| ----------------- | ------------------------------------------------------------------------- |
| `folder_path`     | Directory holding the input file.                                         |
| `file`            | Input file name (with extension). `rio`-readable formats supported.       |
| `encrypted_file`  | Output file name (with extension). Parquet or CSV recommended.            |
| `vars_to_encrypt` | Comma-separated list of columns in the input file to hash.                |
| `vars_to_remove` | Comma-separated list of columns to drop from the output (may be empty).    |
| `to_encrypt`      | `"X"` to include the row in the run, anything else to skip it.            |

Duplicated `encrypted_file` values are automatically disambiguated with
a `DUPL<n>_` prefix, so the output paths stay unique.

## A complete example

The example below uses a 10-row CSV fixture shipped with the package
(`inst/extdata/persons.csv`). Every write happens inside
`tempdir()`, so the example leaves nothing behind on your machine.

```{r paths-setup}
input_file    <- system.file("extdata", "persons.csv", package = "cryptRopen")
input_folder  <- dirname(input_file)

work_dir           <- tempfile("cryptR_vignette_")
dir.create(work_dir)
mask_dir           <- file.path(work_dir, "mask")
output_dir         <- file.path(work_dir, "output")
intermediate_dir   <- file.path(work_dir, "intermediate")
dir.create(mask_dir)
dir.create(output_dir)
dir.create(intermediate_dir)
```

Build a minimal mask: one row, pseudonymising `email` and dropping
`joined_date`.

```{r build-mask}
mask <- data.frame(
  folder_path     = input_folder,
  file            = basename(input_file),
  encrypted_file  = "persons_crypt.csv",
  vars_to_encrypt = "email",
  vars_to_remove  = "joined_date",
  to_encrypt      = "X",
  stringsAsFactors = FALSE
)
writexl::write_xlsx(mask, file.path(mask_dir, "mask.xlsx"))
```

Dispatch the job. `crypt_r()` returns immediately, before the workers
have finished; the value is a `cryptR_job` handle.

```{r crypt-r-dispatch, results = "hide"}
job <- crypt_r(
  mask_folder_path  = mask_dir,
  mask_file         = "mask.xlsx",
  output_path       = output_dir,
  intermediate_path = intermediate_dir,
  encryption_key    = "vignette-key",
  algorithm         = "md5",
  n_workers         = 1L
)
```

```{r job-class}
class(job)
```

## Inspecting the job

Three orthogonal views are available at any time during or after the
run.

### `cryptR_status()` — per-task lifecycle

```{r status}
cryptR_wait(job)
cryptR_status(job)
```

Columns: `encrypted_file`, `state` (`running` / `done` / `failed`),
`error_message`, plus `start_time`, `end_time`, `duration_sec`,
`n_rows_processed`. `NA` until the task resolves.

### `summary(job)` — dashboard

```{r summary}
summary(job)
```

Returns a compact object with task counts by state, elapsed seconds,
active workers, total rows processed, output path, and the full
status data frame under `$status`.

### `cryptR_results()` — disk-oriented view

```{r results}
cryptR_results(job)
```

One row per filtered mask row with the expected `output_file_path`, a
live `exists` flag, and the `size_bytes` / `sha256` the worker
recorded right after writing the output. Joins cleanly with
`cryptR_status()` on `encrypted_file`.

## Finalising the job

`cryptR_collect()` waits for the tasks to resolve (if you have not
called `cryptR_wait()` already), re-publishes the correspondence
tables produced inside the mirai daemons back into your session's
environment, writes the recap log, and tears down any daemons
`crypt_r()` created for the run.

```{r collect}
job <- cryptR_collect(job)
```

The correspondence table is now visible in the parent session:

```{r get-tc}
tcs <- get_correspondence_tables()
names(tcs)
head(tcs$tc_persons_crypt)
```

The recap log is an xlsx file whose name embeds the run timestamp:

```{r log-file}
list.files(output_dir, pattern = "^log_crypt_r_.*\\.xlsx$")
```

An auto-watcher (based on `later::later()`) writes the same log as soon
as the last mirai task resolves, so a manual `cryptR_collect()` call
is not strictly required for the log to appear — it just becomes a
reliable synchronisation point.

## Options recap

| Parameter              | Default            | Purpose                                                                 |
| ---------------------- | ------------------ | ----------------------------------------------------------------------- |
| `algorithm`            | `"md5"`            | Hashing algorithm. Anything `digest::digest()` accepts.                 |
| `correspondence_table` | `TRUE`             | Produce the TC on disk (parquet) and in `.cryptRopen_env`.              |
| `engine`               | `"auto"`           | `"auto"`, `"in_memory"`, or `"streaming"`. See routing table below.     |
| `chunk_size`           | `1e6`              | Rows per chunk when the selected engine is streaming. Ignored otherwise. |
| `n_workers`            | `NULL`             | Mirai daemons to spawn. Default heuristic: `min(detectCores()-1, n_rows, 8L)`. |

### Engine routing

| `engine`       | Input / output combination        | Effective engine                     |
| -------------- | --------------------------------- | ------------------------------------ |
| `"in_memory"`  | any                               | `.process_mask_row_in_memory()`      |
| `"auto"` / `"streaming"` | parquet-in + parquet-out | `.process_mask_row_streaming()`      |
| `"auto"` / `"streaming"` | csv-in + csv-out         | `.process_mask_row_csv_streaming()`  |
| `"auto"` / `"streaming"` | mixed, rds, xlsx, …      | fallback to `in_memory`              |

The streaming engines read with an `arrow` Scanner by chunks and write
incrementally, keeping memory usage bounded regardless of input size.

### Daemons ownership

If no mirai daemons are active on the default profile when `crypt_r()`
is called, it spawns `n_workers` of its own and flags the job for
automatic teardown by `cryptR_collect()`. If daemons are already
running, `crypt_r()` reuses them and `cryptR_collect()` leaves them
alone — useful when you wrap several `crypt_r()` calls inside a
broader parallel session you manage yourself.

### Error handling

A failure on one mask row never interrupts the others. The failure is
captured on the corresponding mirai task and surfaces as:

* `state == "failed"` in `cryptR_status()`,
* a non-`NA` `error_message` column,
* `success = FALSE` in the xlsx recap log.

Invalid `output_path` or `intermediate_path` — paths that don't exist,
or aren't scalar non-empty character — are caught before any task is
dispatched, with a clear error.

## Cleanup

Nothing was written outside `tempdir()`, so the example cleans itself
up when the R session ends. You can also drop the folder explicitly:

```{r cleanup}
unlink(work_dir, recursive = TRUE)
```

## See also

* `?crypt_r`, `?cryptR_status`, `?cryptR_wait`, `?cryptR_collect`,
  `?cryptR_results`, `?summary.cryptR_job`.
* `vignette("cryptRopen")` for the in-session workflow with
  `crypt_data()`.
* `NEWS.md` for the full change history.
