---
title: "Getting started with cryptRopen"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with cryptRopen}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## What cryptRopen is for

`cryptRopen` pseudonymises variables inside data frames so that sensitive
values (identifiers, emails, names) can be shared or stored in a
derivative form without exposing the originals. The pseudonymisation is
**deterministic**: the same input value paired with the same key always
produces the same output, so joins across tables remain possible on the
pseudonymised columns. A side table — the *correspondence table* —
keeps the mapping between original and pseudonymised values for the
party that holds the key.

Typical use cases: preparing an extract for an analytics team that
should not see customer identifiers; sharing a dataset with an external
partner; feeding a downstream pipeline while keeping the original
identifiers under control.

## Pseudonymisation, not encryption

The `crypt` in the package name is historical. The underlying
transformation is a **salted hash**, not reversible encryption:

* A normalised version of each input value is concatenated with a
  user-provided salt (`key`) and hashed with `digest::digest()`.
* The default algorithm is `"md5"`. Any algorithm supported by
  `digest::digest()` (`"sha1"`, `"sha256"`, `"sha512"`, …) is accepted
  via the `algorithm` parameter. MD5 collisions are theoretically
  possible but have negligible impact at typical dataset scales; pick a
  stronger algorithm if your threat model calls for it.
* Holding the output does **not** let anyone recover the input. Only
  the correspondence table (or a rainbow-table-style attack on a weak
  key) does.

Consequence: cryptRopen is suitable for pseudonymisation under
regulations such as GDPR, not for confidentiality-bearing encryption.

## Installation

```{r install, eval = FALSE}
# install.packages("pak")
pak::pak("danielrak/cryptRopen")
```

```{r setup}
library(cryptRopen)
```

## Hashing a single vector

`crypt_vector()` is the lowest-level entry point. It takes a vector, a
key, and a hash algorithm, and returns a character vector of the same
length. `NA` and empty strings stay `NA`. Input is uppercased, trimmed,
and stripped of internal spaces before hashing, so `"Alice"`,
`"alice "` and `" ALICE "` collapse to the same hash.

```{r crypt-vector-basic}
crypt_vector(
  vector = c("Alice", "Bob", "Charlie"),
  key = "my-secret-salt",
  algo = "md5"
)
```

Determinism and the NA / empty-string rules:

```{r crypt-vector-properties}
# Same value + same key => same hash, every time.
identical(
  crypt_vector("Alice", key = "k", algo = "md5"),
  crypt_vector("alice ", key = "k", algo = "md5")
)

# NAs and empty strings are preserved.
crypt_vector(c("Alice", NA, "", "  "), key = "k", algo = "md5")
```

## Hashing one or several columns of a data frame

`crypt_data()` operates on a data frame and returns a new data frame
with the requested columns replaced by their hashed counterparts
(suffix `_crypt`). Original columns can optionally be dropped
(`vars_to_remove`).

```{r crypt-data-basic}
head(mtcars, 3)

encrypted <- crypt_data(
  loaded_dataset            = head(mtcars, 3),
  vars_to_encrypt           = "mpg",
  vars_to_remove            = "cyl",
  encryption_key            = "demo-key",
  algorithm                 = "md5",
  correspondence_table      = TRUE,
  correspondence_table_label = "demo"
)
encrypted
```

The correspondence table is **not** returned alongside the data frame;
it is stored inside a package-private environment and retrieved via
`get_correspondence_tables()`:

```{r get-tc}
tcs <- get_correspondence_tables()
names(tcs)
tcs$tc_crypt_demo
```

Tables accumulate across calls in the same session. Pass a character
vector to `names` to pull only the ones you need:

```{r get-tc-filter}
get_correspondence_tables(names = "tc_crypt_demo")
```

Pseudonymising multiple columns at once is a matter of extending
`vars_to_encrypt`:

```{r crypt-data-multi}
crypt_data(
  loaded_dataset            = head(mtcars, 3),
  vars_to_encrypt           = c("mpg", "hp"),
  encryption_key            = "demo-key",
  correspondence_table      = TRUE,
  correspondence_table_label = "demo_multi"
)
```

## When you need to batch multiple files

`crypt_data()` is the right tool when you hold one data frame in memory
and want an encrypted version of it. For an industrialised workflow —
several input files specified in an Excel mask, outputs written to
disk, parallel execution, a recap log — see `vignette("crypt_r-workflow")`.

## Related resources

* `?crypt_vector`, `?crypt_data`, `?crypt_r` — the three public
  entry points.
* `?get_correspondence_tables` — retrieval of the correspondence
  tables stored in the package-private environment.
* `NEWS.md` — full change history.
